---
title: "A Series of Bayesian Fish Growth Models"
author: "Alexandra Norelli"
date: "2024-09-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(warning = FALSE)
```

## Clean and Examine White Perch Otolith Data

```{r, include = FALSE}
# Load packages
library(tidyverse)
library(FSA)

# Read in the white perch age data (wtpa)
wtpa <- read.csv("all_bc.csv")

# Reformat the back-calculated length at ages
bc_wtpa <- pivot_longer(wtpa, c(ONE.L, TWO.L, THREE.L, FOUR.L, FIVE.L, SIX.L, SEVEN.L, EIGHT.L), names_to = "otolith", values_drop_na = TRUE)
bc_wtpa$otolith <-as.factor(bc_wtpa$otolith)
bc_wtpa$otolith <- factor(bc_wtpa$otolith, levels = c("ONE.L", "TWO.L", "THREE.L", "FOUR.L", "FIVE.L", "SIX.L", "SEVEN.L", "EIGHT.L"))
bc_wtpa$otolith <- as.numeric(bc_wtpa$otolith)

# Add a type of measurement variable
wtpa$type <- rep("Measured")
bc_wtpa$type <- rep("Back-Calc")

# Merge the two types of datasets with cleaned variables
cleaned_wtpa <- data.frame("Length" = c(bc_wtpa$value, wtpa$Length), "Age" = c(bc_wtpa$otolith, wtpa$Age), "Site" = c(bc_wtpa$Site, wtpa$Site), "Type" = c(bc_wtpa$type, wtpa$type))
cleaned_wtpa$Type <- as.factor(cleaned_wtpa$Type)

```

## Histograms and Scatterplots

``` {r}
# Histogram of Fish Length
ggplot(cleaned_wtpa, aes(Length, fill = Type)) +
  geom_histogram(binwidth = 10, color = "black") + scale_fill_manual(values=c("white", "black")) +ggtitle("Histogram of Fish Length (cm)") + ylab("Count")+ theme_classic()

# Histogram of Fish Ages
ggplot(cleaned_wtpa, aes(Age, fill = Type)) +
  geom_histogram(binwidth = 1, color = "black") + scale_fill_manual(values=c("white", "black"))+ggtitle("Histogram of Fish Age") + ylab("Count") + theme_classic()

```

``` {r}
# Scatterplot of raw length at age by type
ggplot(cleaned_wtpa, aes(x = Age, y = Length, color = Type)) +
  geom_point() + 
  scale_color_manual(values=c("#999999", "black")) + 
  theme_bw() + 
  ggtitle("Raw length at ages by measurement type")

# Measured-type length at age by location
mes <- filter(cleaned_wtpa, Type == "Measured")
ggplot(mes, aes(x = Age, y = Length)) +
  geom_point(aes(shape = Site, color = Site)) +
  theme_bw() +
  ggtitle("Measured otolith length at ages by location")

# All length at ages by location
ggplot(cleaned_wtpa, aes(x = Age, y = Length)) +
  geom_point(aes(shape = Site)) + 
  theme_bw() + 
  ggtitle("All length at ages by location")

```


## JAGS Models

These chunks create and run 3 JAGS models, one with just the measured length at ages, one with all the length at ages, and a third with a subset of ages 1-5.  Loosely based off Ortiz de Zárate and Babcock (2016).  

```{r, include = FALSE}
# Load Packages
library(R2jags)
library(ggmcmc)
```

### Model 1: Measured Otoliths Growth Curve

```{r}

##### All Measured Data ######

# Create model.file for JAGS
# I looped through 282 observations to find a von Bertalanffy curve that would fit the measured data
# The priors were selected based off the general range of White Perch growth curves
write("model
      {
      for(i in 1:282){
        PredL[i] <- Linf * (1 - exp(-K*(Age[i] - Age0)))
        logPredL[i] <- log(PredL[i])
        Length[i] ~ dlnorm(logPredL[i], tau)
			  logObsL[i] <-log(Length[i]) 	     # log transfomration of observed value
			  resid[i] <- logObsL[i]-logPredL[i]  # residuals
		 	  Rep[i] ~ dlnorm(logPredL[i], tau) # replicated data set
		 	  logRep[i] <-log(Rep[i]) # replicated data set
		 	  Prob[i] <- step(logRep[i] - logObsL[i])  # Probability replicated data>real data
        sresid2[i]<-(logObsL[i]-logPredL[i])*(logObsL[i]-logPredL[i])*tau   # stand. residuals squared
        rep.sresid2[i]<-(logRep[i]-logPredL[i])*(logRep[i]-logPredL[i])*tau  # replicated sresid squared
        sstest[i] <- (logObsL[i]-logMean)*(logObsL[i]-logMean)*tau 
      }
      #priors specification
      K ~ dunif(0,2)
      Age0 ~ dunif(-10,0)
      Linf ~ dunif(200,400)
      tau ~ dgamma(0.001,0.001)
      
      #Derived parameters
      logMean <-mean(logObsL[])
      mean.prob<-mean(Prob[])
      chi.square.obs<-sum(sresid2[])
      chi.square.rep<-sum(rep.sresid2[])
      sstot <-sum(sstest[])
      p.value<-step(chi.square.obs-chi.square.rep)
      r.squared <- 1-(chi.square.obs/sstot)
      r.squared2 <-1-1/(tau*sd(logObsL[])*sd(logObsL[]))
      
      }", file = "vonBwtp.txt")

# Set initial values for JAGS
init1 <- list(tau=1,K=0.6)
init2 <- list(tau=0.5,K=0.4)

# Create a copy of wtpa
wtp <- wtpa

# Run JAGS model with 400,000 iterations, 50,000 burn in, and 100 thin
wtpjags=jags(wtp,list(init1,init2),model.file="vonBwtp.txt",
             parameters.to.save=c("K","Linf","tau","Age0","p.value","r.squared","r.squared2",
                                  "PredL","logPredL"),
             n.chains=2,n.iter=400000,n.burnin=50000,n.thin=100)

# Create and look at JAGS output
res1 <- wtpjags$BUGSoutput
round(res1$summary[c("Age0", "K","Linf","p.value","r.squared","r.squared2"),c("mean","sd","2.5%","50%","97.5%","Rhat","n.eff")],3)

```

``` {r}
# Plot the glm line of the mean with confidence intervals over a scatter plot of the data 
df <- data.frame(res1$summary[paste0("PredL[",1:282,"]"),c("mean","2.5%","50%","97.5%")])
names(df) <- c("mean","lci","median","uci")
df <- cbind(df,wtp)
ggplot(df, aes(x=Age, y=Length, ymin=lci, ymax=uci))+
  geom_point() +
  geom_ribbon(alpha=0.3) +
  theme_bw() +
  geom_line(aes(x=Age,y=mean)) +
  ylab("Length") +
  ggtitle("A von Bertalanffy growth curve of measured white perch otoliths")

```

```{r}
# Take a subset of the jags runs to make gelman plots
mcmc1 <- as.mcmc(wtpjags)
params_of_interest <- c("K","Linf","Age0", "deviance")
subset_mcmc1 <- mcmc1[, params_of_interest, drop = FALSE]
gelman.plot(subset_mcmc1)
```

### Model 2: Measured Otoliths and Back-Calculated Lengths Growth Curve

```{r}
##### All Measured Data and Back-Calculations #####
write("model
      {
      for(i in 1:1368){
        PredL[i] <- Linf * (1 - exp(-K*(Age[i] - Age0)))
        logPredL[i] <- log(PredL[i])
        Length[i] ~ dlnorm(logPredL[i], tau)
        			  logObsL[i] <-log(Length[i]) 	     # log transfomration of observed value
			  resid[i] <- logObsL[i]-logPredL[i]  # residuals
		 	  Rep[i] ~ dlnorm(logPredL[i], tau) # replicated data set
		 	  logRep[i] <-log(Rep[i]) # replicated data set
		 	  Prob[i] <- step(logRep[i] - logObsL[i])  # Probability replicated data>real data
        sresid2[i]<-(logObsL[i]-logPredL[i])*(logObsL[i]-logPredL[i])*tau   # stand. residuals squared
        rep.sresid2[i]<-(logRep[i]-logPredL[i])*(logRep[i]-logPredL[i])*tau  # replicated sresid squared
        sstest[i] <- (logObsL[i]-logMean)*(logObsL[i]-logMean)*tau 
      }
      #priors specification
      K ~ dunif(0,2)
      Age0 ~ dunif(-10,0)
      Linf ~ dunif(200,400)
      tau ~ dgamma(0.001,0.001)
      
      #Derived parameters
      logMean <-mean(logObsL[])
      mean.prob<-mean(Prob[])
      chi.square.obs<-sum(sresid2[])
      chi.square.rep<-sum(rep.sresid2[])
      sstot <-sum(sstest[])
      p.value<-step(chi.square.obs-chi.square.rep)
      r.squared <- 1-(chi.square.obs/sstot)
      r.squared2 <-1-1/(tau*sd(logObsL[])*sd(logObsL[]))
      }", file = "vonBwtp2.txt")

# make a copy of cleaned_wtpa for this model
wtp <- cleaned_wtpa

# run the model
wtpjags3=jags(wtp,list(init1,init2),model.file="vonBwtp2.txt",
             parameters.to.save=c("K","Linf","tau", "Age0","r.squared", "p.value","r.squared2",
                                  "PredL","logPredL"),
             n.chains=2,n.iter=400000,n.burnin=50000,n.thin=100)

# view model outputs
res2=wtpjags3$BUGSoutput
round(res2$summary[c("Age0", "K","Linf","p.value","r.squared","r.squared2"),c("mean","sd","2.5%","50%","97.5%","Rhat","n.eff")],5)

```

```{r}
# select some of the results and plot them
df2<-data.frame(res2$summary[paste0("PredL[",1:1368,"]"),c("mean","2.5%","50%","97.5%")])
names(df2)<-c("mean","lci","median","uci")
df2<-cbind(df2, wtp)
ggplot(df2, aes(x=Age, y=Length, ymin=lci, ymax=uci)) +
  geom_point() +
  geom_ribbon(alpha=0.3) +
  theme_bw() +
  geom_line(aes(x=Age,y=mean)) +
  ylab("Length") +
  ggtitle("A von Bertalanffy growth curve of all types of white perch otoliths")


```

```{r}
# Take a subset of the jags runs to make gelman plots
mcmc2 <- as.mcmc(wtpjags3)
subset_mcmc2 <- mcmc2[, params_of_interest, drop = FALSE]
gelman.plot(subset_mcmc2)
```

### Model 3: Modified Sample of Ages 1-5 Growth Curve

```{r}
##### Subset #####
# The final model used a subsample of ages 1-5 where each age had 100 samples where the measured length at ages were supplemented by however many back-calculated ages were needed to reach 100 samples.   


# take the subset and make every age group n = 100.  
df_3 = 0
for(i in 1:5){
  #only get less than or equal to age i
  s <- filter(cleaned_wtpa, Age <= i)
  s <- filter(s, (i-1) < Age)
  #only get = to age i 
  s.b <- filter(s, Type == "Back-Calc")
  s.a <- filter(s, Type == "Measured")
  x = 100 - nrow(s.a)
  s.b <- sample_n(s.b, x)
  df_3 <- rbind(df_3, s.a, s.b)
}
df_3 <- df_3[-1,]

df_3$round <- ceiling(df_3$Age)

# plot the new age histograms
ggplot(df_3, aes(round, fill = Type)) +
  geom_histogram(bins = 5, color = "black") +
  scale_fill_manual(values=c("white", "black")) +
  ggtitle("Histogram of Fish Age") + 
  ylab("Count") + 
  theme_classic()

```

```{r}
# The subsetted model for jags
write("model
      {
      for(i in 1:500){
        PredL[i] <- Linf * (1 - exp(-K*(Age[i] - Age0)))
        logPredL[i] <- log(PredL[i])
        Length[i] ~ dlnorm(logPredL[i], tau)
        logObsL[i] <-log(Length[i]) 	     # log transfomration of observed value
			  resid[i] <- logObsL[i]-logPredL[i]  # residuals
		 	  Rep[i] ~ dlnorm(logPredL[i], tau) # replicated data set
		 	  logRep[i] <-log(Rep[i]) # replicated data set
		 	  Prob[i] <- step(logRep[i] - logObsL[i])  # Probability replicated data>real data
        sresid2[i]<-(logObsL[i]-logPredL[i])*(logObsL[i]-logPredL[i])*tau   # stand. residuals squared
        rep.sresid2[i]<-(logRep[i]-logPredL[i])*(logRep[i]-logPredL[i])*tau  # replicated sresid squared
        sstest[i] <- (logObsL[i]-logMean)*(logObsL[i]-logMean)*tau 
      }
      #priors specification
      K ~ dunif(0,2)
      Age0 ~ dunif(-10,0)
      Linf ~ dunif(200,400)
      tau ~ dgamma(0.001,0.001)
      
      #Derived parameters
      logMean <-mean(logObsL[])
      mean.prob<-mean(Prob[])
      chi.square.obs<-sum(sresid2[])
      chi.square.rep<-sum(rep.sresid2[])
      sstot <-sum(sstest[])
      p.value<-step(chi.square.obs-chi.square.rep)
      r.squared <- 1-(chi.square.obs/sstot)
      r.squared2 <-1-1/(tau*sd(logObsL[])*sd(logObsL[]))
      
      }", file = "vonBwtp3.txt")

# create a copy of the data generated for the subset model
wtp <- df_3

# run the JAGS model
wtpjags5=jags(wtp,list(init1,init2),model.file="vonBwtp3.txt",
              parameters.to.save=c("K","Linf","tau", "Age0","r.squared", "p.value","r.squared2",
                                   "PredL","logPredL"),
              n.chains=2,n.iter=400000,n.burnin=50000,n.thin=100)

# Print out the outputs
res3=wtpjags5$BUGSoutput
round(res3$summary[c("Age0", "K","Linf","p.value","r.squared","r.squared2"),c("mean","sd","2.5%","50%","97.5%","Rhat","n.eff")],5)

```


```{r}
# subset the data and create a plot of the results
df3<-data.frame(res3$summary[paste0("PredL[",1:500,"]"),c("mean","2.5%","50%","97.5%")])
names(df3)<-c("mean","lci","median","uci")
df3<-cbind(df3,wtp)
ggplot(df3,aes(x=Age, y=Length, ymin=lci, ymax=uci)) +
  geom_point() +
  geom_ribbon(alpha=0.3) +
  theme_bw() +
  geom_line(aes(x=Age,y=mean)) +
  ylab("Length") +
  ggtitle("A von Bertalanffy growth curve of sub-sampled age 1-5 white perch otoliths")

```

```{r}
# Take a subset of the jags runs to make gelman plots
mcmc3 <- as.mcmc(wtpjags5)
subset_mcmc3 <- mcmc3[, params_of_interest, drop = FALSE]
gelman.plot(subset_mcmc3)
```

## Conclusions

### Model 1: Measured Density Plots

```{r}
# Create density plots of the 3 model runs
par(mfrow = c(2,2))
densplot(subset_mcmc1)   
```

### Model 2: All Data Density Plots

```{r}
par(mfrow = c(2,2))
densplot(subset_mcmc2)   
```

### Model 3: Sub-Sample Density Plots

```{r}
par(mfrow = c(2,2))
densplot(subset_mcmc3)   
```

```{r}
# merge the 3 model datasets for plotting
df <- df %>% 
  mutate(df, model = "M1:Measured") %>%
  select(model, Age, mean)
df2 <- df2 %>% 
  mutate(df2, model = "M2:All Data") %>%
  select(model, Age, mean)
df3 <- df3 %>% 
  mutate(df3, model = "M3:Sub-Sample") %>%
  select(model, Age, mean)
glm_results <- rbind(df, df2, df3)

# create the final comparison plot
ggplot(cleaned_wtpa, aes(x=Age, y=Length)) +
  geom_point() +
  theme_bw() +
  geom_line(data = glm_results, aes(x=Age, y=mean, color = model)) +
  ylab("Length") +
  ggtitle("A comparison of three von Bertalanffy growth curves for white perch")

```